Left-Linear Rewriting in Adhesive Categories

### 2024-07-08 concur/final.tex
minima formattazione

Left-linear (w.r.t.~$\reg(\cat{EGraphs})$) rewriting systems on $\cat{EGraphs}$ where the right half of a rule might be mono but not regular 
are well-switching. They have been used in~\cite{BaldanGM06} for the graphical implementation of nominal calculi, where,
differently from~\cite{Gad07}, as a result of name passing the received name is not merged with a local one at the receiver,
but put instead in the same equivalence class, thus better tracing the causal dependencies among processes reductions.
\end{example}

@@

\noindent
\begin{minipage}{.8\textwidth}
  \setlength\parindent{1.5em}
  Instead, if in $\dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2$ we have that $\dder{D}_1$ and $\dder{D}_2$ are
independent, and we switch $\dder{D}_0$ to the last position, obtaining a
derivation $\dder{D}_1'\cdot \dder{D}_2'\cdot  \dder{D}_0''$, as in the diagram on the right, the steps $\dder{D}_1'$ and $\dder{D}_2'$ could be no
longer independent, unless an additional condition is
satisfied. Intuitively, this is due to the fact that dealing with
left-linear rewriting systems introduces disjunctive forms of
causality. Hence, if in $\dder{D}_0\cdot \dder{D}_1\cdot \dder{D}_2$ we have that $\dder{D}_0$ and $\dder{D}_1$ are
disjunctive causes of $\dder{D}_2$, at least one of the two is needed to
enable the application of $\dder{D}_2$ and this explain why in
$\dder{D}_1'\cdot \dder{D}_2'\cdot \dder{D}_0''$ we have that $\dder{D}_1'$ and $\dder{D}_2'$ are no longer
independent. This is exemplified below.
\end{minipage}
\begin{minipage}{.18\textwidth}
\ \hspace{-14mm}
\begin{tikzpicture}[node/.style={draw, minimum size=8mm}]
     \begin{tikzcd}[row sep=2mm, column sep=1mm]
      \dder{D}_0 \arrow[perm,dr]  & \dder{D}_1 \arrow[perm,dl] & \dder{D}_2 \arrow[perm,d]\\
      \dder{D}_1' \arrow[perm,d]  & \dder{D}_0' \arrow[perm,dr] & \dder{D}_2 \arrow[perm,dl]\\
      \dder{D}_1'   & \dder{D}_2' & \dder{D}_0''
    \end{tikzcd}
  \end{tikzpicture} 
\end{minipage}

\begin{example}
  \label{ex:non-global}
  Consider the rewriting system in $\cat{Graph}$ consisting of the
  rules $\lambda_0$, $\lambda_1$, and $\lambda_2$ in
  Fig.~\ref{fi:rules-cas}.
  %\todo{P: rev.2 dice ``In the top-left and
   % top-middle nodes of $\rho_3$, the labels "12" should be "1" and
   % "2".''. Non e' corretto, ma magari si capisce poco ...}  
   Consider the derivation $\der{F}$ in Fig.~\ref{fi:derF}. Note that rule
  $\lambda_2$ needs the black self-loop on node $12$ to be applied. This can be
  generated either by $\lambda_0$ or by $\lambda_1$ merging nodes $1$ and
  $2$, hence at least one of them must precede the application of
  $\lambda_2$. Indeed, in $\der{F}$ it is easily seen that the second
  and the third step applying $\lambda_1$ and $\lambda_2$ are
  independent. However, we could switch twice the application of
  $\lambda_0$, bringing it in the last position, obtaining a derivation
  $\der{F}'$ as depicted in Fig.~\ref{fi:derF1}, where the
  applications of $\lambda_1$ and $\lambda_2$ are no longer independent.

  \begin{figure}[t]
    \input{Figure/rules-cas.tex}    
    \caption{A new rewriting system in $\cat{Graph}$.}
    \label{fi:rules-cas}
  \end{figure}

  \begin{figure}[t]
    \input{Figure/derF.tex}    
    \caption{The derivation $\der{F}$.} 
    \label{fi:derF}
  \end{figure}

  \begin{figure}[t]
    \input{Figure/derF1.tex}    
    \caption{The derivation $\der{F}'$.} 
    \label{fi:derF1}
  \end{figure}
\end{example}

\begin{restatable}[Globality of independence]{lemma}{lemIndepGlobalLeft}
  \label{lem:indep-global-left}
  Let $\X$ be an $\mathcal{M}$-adhesive category and $(\X,\R)$ a
  well-switching rewriting system.  Let
  $\dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2$ be a three-steps
  derivation.
  \begin{enumerate}
  \item
    \label{lem:indep-global-left:1}
    If $\dder{D}_0$ and $\dder{D}_1$ are switchable and there is a
    switching sequence
    $\dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2 \shift{(1,2)}
    \dder{D}_0 \cdot \dder{D}_2' \cdot \dder{D}_1' \shift{(0,1)}
    \dder{D}_2'' \cdot \dder{D}_0' \cdot \dder{D}_1'$ then
    in the last derivation $\dder{D}_0'$ and $\dder{D}_1'$ are switchable;

@@

\full{[Proof~\ref{lemIndepGlobalLeft-proof}]}

The second lemma captures the essence of the consistency of
switchings. It shows that if we have a derivation consisting of three
rewriting steps, say $\dder{D}_0\cdot \dder{D}_1\cdot  \dder{D}_2$ and we invert them leading to a
sequence reordered as $\dder{D}_2\cdot  \dder{D}_1\cdot  \dder{D}_0$, we obtain the same resulting derivation,
independently from the order of the switchings.

@@

Still, we can identify a
canonical form for switching derivations: at each step one can apply a switch to the inversion with largest index.

>>>>

Left-linear rewriting systems %(with respect to~$\reg(\cat{EGraphs})$) 
with respect to $\reg(\cat{EGraphs})$ are well-switching. They have been used in~\cite{BaldanGM06} for the graphical implementation of nominal calculi, where,
differently from~\cite{Gad07}, as a result of name passing the received name is not merged with a local one, % at the receiver,
but put in the same equivalence class, better tracing the causal dependencies among reductions.
\end{example}

@@

\noindent
\begin{minipage}{.8\textwidth}
%  \setlength\parindent{1.5em}
  If instead $\dder{D}_1$ and $\dder{D}_2$ are
independent in $\dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2$ and we switch $\dder{D}_0$ to the last position, obtaining a
derivation $\dder{D}_1'\cdot \dder{D}_2'\cdot  \dder{D}_0''$, as in the diagram on the right, the steps $\dder{D}_1'$ and $\dder{D}_2'$ could be no
longer independent. %unless an additional condition is satisfied. 
Intuitively, this is due to the fact that dealing with
left-linear rewriting systems introduces disjunctive forms of
causality. Hence, if in $\dder{D}_0\cdot \dder{D}_1\cdot \dder{D}_2$ we have that $\dder{D}_0$ and $\dder{D}_1$ are
disjunctive causes of $\dder{D}_2$, at least one of the two is needed to
enable the application of $\dder{D}_2$ and this explain why in
$\dder{D}_1'\cdot \dder{D}_2'\cdot \dder{D}_0''$ we have that $\dder{D}_1'$ and $\dder{D}_2'$ are no longer
independent. This is exemplified below.
\end{minipage}
\begin{minipage}{.18\textwidth}
\ \hspace{-14mm}
\begin{tikzpicture}[node/.style={draw, minimum size=8mm}]
     \begin{tikzcd}[row sep=2mm, column sep=1mm]
      \dder{D}_0 \arrow[perm,dr]  & \dder{D}_1 \arrow[perm,dl] & \dder{D}_2 \arrow[perm,d]\\
      \dder{D}_1' \arrow[perm,d]  & \dder{D}_0' \arrow[perm,dr] & \dder{D}_2 \arrow[perm,dl]\\
      \dder{D}_1'   & \dder{D}_2' & \dder{D}_0''
    \end{tikzcd}
  \end{tikzpicture} 
\end{minipage}

\begin{example}
  \label{ex:non-global}
  Consider the rewriting system in $\cat{Graph}$ consisting of the
  rules $\lambda_0$, $\lambda_1$, and $\lambda_2$ in
  Fig.~\ref{fi:rules-cas}.
  %\todo{P: rev.2 dice ``In the top-left and
   % top-middle nodes of $\rho_3$, the labels "12" should be "1" and
   % "2".''. Non e' corretto, ma magari si capisce poco ...}  
   Consider the derivation $\der{F}$ in Fig.~\ref{fi:derF}. Note that rule
  $\lambda_2$ needs the black self-loop on node $12$ to be applied. This can be
  generated either by $\lambda_0$ or by $\lambda_1$ merging nodes $1$ and
  $2$, hence at least one of them must precede the application of
  $\lambda_2$. Indeed, in $\der{F}$ it is easily seen that the second
  and the third step applying $\lambda_1$ and $\lambda_2$ are
  independent. However, we could switch twice the application of
  $\lambda_0$, bringing it in the last position, obtaining a derivation
  $\der{F}'$ as depicted in Fig.~\ref{fi:derF1}, where the
  applications of $\lambda_1$ and $\lambda_2$ are no longer independent.
   %
  \begin{figure}[t]
    \input{Figure/rules-cas.tex}    
    \caption{A new rewriting system in $\cat{Graph}$.}
    \label{fi:rules-cas}
  \end{figure}
%
  \begin{figure}[t]
    \input{Figure/derF.tex}    
    \caption{The derivation $\der{F}$.} 
    \label{fi:derF}
  \end{figure}
%
  \begin{figure}[t]
    \input{Figure/derF1.tex}    
    \caption{The derivation $\der{F}'$.} 
    \label{fi:derF1}
  \end{figure}
\end{example}
%
\begin{restatable}[Globality of independence]{lemma}{lemIndepGlobalLeft}
  \label{lem:indep-global-left}
  Let $\X$ be an $\mathcal{M}$-adhesive category and $(\X,\R)$ a
  well-switching rewriting system.  Let
  $\dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2$ be a three-steps
  derivation.
  \begin{enumerate}
  \item
    \label{lem:indep-global-left:1}
    If $\dder{D}_0$ and $\dder{D}_1$ are switchable and there is a
    switching sequence
    $\dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2 \shift{(1,2)}
    \dder{D}_0 \cdot \dder{D}_2' \cdot \dder{D}_1' \shift{(0,1)}
    \dder{D}_2'' \cdot \dder{D}_0' \cdot \dder{D}_1'$ then
    in the last derivation $\dder{D}_0'$ and $\dder{D}_1'$ are switchable;

@@

%\full{[Proof~\ref{lemIndepGlobalLeft-proof}]}
%
The second lemma captures the essence of the consistency of
switchings. If in a derivation of three
rewriting steps, say $\dder{D}_0\cdot \dder{D}_1\cdot  \dder{D}_2$, we invert them leading to a
sequence reordered as $\dder{D}_2\cdot  \dder{D}_1\cdot  \dder{D}_0$, we obtain the same derivation,
independently from the order of switchings.

@@

\newpage
Still, we can identify a
canonical form for switching derivations: at each step one can apply a switch to the inversion with largest index.
